\documentclass{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{natbib}

\title{Mechanistic Interpretability of Reasoning in LLMs via Sparse Autoencoders: A Phased, Resource-Efficient, Falsifiable Framework}
\author{Benjamin Huh}
\date{February 2026}

\begin{document}

\maketitle

\begin{abstract}
This document outlines a comprehensive, phased framework to evaluate sparse autoencoders (SAEs) for extracting human-interpretable, causally verifiable reasoning features from large language model (LLM) activations. Building on SAE successes~\citep{AnthropicMappingMind} and limitations~\citep{SAE-Identifiability,SAE-Random,SAE-Empirical}, we progress from ground-truth systems to LLMs, with strict falsification criteria. Key innovations include decorrelation losses, probe-guided training with leakage diagnostics, staged alignment, and causal attribution protocols. Resource optimizations enable execution on single high-end GPUs (e.g., RTX 6000). Full details cover theory, architecture, datasets, workflow, validation, risks, and scaling.
\end{abstract}

\section{Introduction and Motivation}
LLMs excel on reasoning benchmarks, yet internal multi-step processes (e.g., hypothesis formation, constraint propagation) remain opaque. Chain-of-thought (CoT) is unreliable~\citep{Shu2025SurveySAE}. SAEs address this by learning sparse, monosemantic latents via dictionary learning~\citep{AnthropicMappingMind,ScalingMonosemanticityBlog}.

\textbf{Research Question:} Can SAEs recover stable, causal reasoning primitives across phases, or do they reflect artifacts?

Phased to falsify: Failures (e.g., no causal alignment) halt progression~\citep{SAE-Identifiability,SAE-Random}.

\section{Theory and Background}
\subsection{Sparse Autoencoders}
SAE: Encoder $f$, decoder $g$ s.t. $X \approx g(f(X))$, $L = \|X - \hat{X}\|_2^2 + \lambda \|h\|_1 + R(h)$, $h = f(X)$~\citep{Shu2025SurveySAE}. Overcomplete bases yield monosemanticity, outperforming neurons~\citep{AnthropicMappingMind,SAE-Monosemantic}.

Limitations: Non-identifiable without sparsity~\citep{SAE-Identifiability}; random nets interpretable~\citep{SAE-Random}; domain-sensitive~\citep{SAE-Domain,SAE-Empirical}.

\subsection{Monosemanticity}
Features activate consistently for one concept; polysemy fragments axes~\citep{Zhu2025AbsTopK}.

\section{Experimental Formulation}
\subsection{Activations and Hooks}
Post-MLP residual + MLP hidden (nonlinear updates). Expand to mid-late layers if needed. Token-level from reasoning datasets.

Efficiency: On-the-fly streaming, fp16, 1-2 layers, small models (1B-7B).

\subsection{SAE Architecture}
Overcomplete latent ($k=4$-$8 \times d$), ReLU. Decoder reconstructs $X$.

\subsection{Loss Function}
\[
L_\text{total} = \|X - \hat{X}\|_2^2 + \lambda \|h\|_1 + \beta \sum_{i<j} w_i^\top w_j + \gamma L_\text{probe} + \delta L_\text{temporal}.
\]
Decorrelation: Penalizes redundancy (novel directions, shared subspaces OK). Temporal optional: $\|h_t - h_{t+1}\|_2^2$. Probes: Hypothesis/constraint tasks; ablation if gap $>$5\%.

\section{Datasets and Tasks}
Structured reasoning: Multi-step math, logic, Sudoku. Corpora: GSM8K, CoT-Collection, OpenR1-Math-220k, Reasoning Traces, REVEAL, TRIP, WIQA. Subsample for efficiency.

\section{Phased Framework}
\begin{enumerate}
\item \textbf{Ground-Truth (Phase 1)}: BFS/DFS, stacks; align SAE to states~\citep{SAE-Monosemantic}.
\item \textbf{Synthetic Transformers (Phase 2)}: Grid/logic; causal perturbations.
\item \textbf{Controlled CoT (Phase 3)}: Labeled steps; probes; 100\% accuracy validation.
\item \textbf{Frontier LLMs (Phase 4)}: Multi-model analysis (GPT-2-medium, Pythia-1.4B, Gemma-2B, Phi-2); feature sparsity metrics.
\item \textbf{Interpretability Validation (Phase 4B)}: Feature purity, selectivity analysis, semantic descriptions.
\item \textbf{Causal Attribution (Phase 5)}: Validate importance estimates; rank features by true causal impact; identify model-architecture dependencies.
\item \textbf{CoT-SAE Extension (Phase 6, proposed)}: Step-level causal testing; reasoning controllability; cross-model universality.
\end{enumerate}

\section{Experimental Workflow}
\begin{enumerate}
\item Capture (stream, monitor).
\item Train Phase 1: Recon + sparsity.
\item Phase 2: +Decorr/probes.
\item Align: Multi-heuristic (regex, similarity), confidence-weighted; EM refine.
\item Inspect: LM descriptions, clustering.
\item Validate: Purity, causal ($h \leftarrow h + \epsilon e_i$).
\item Iterate.
\end{enumerate}

\section{Interpretability Protocols}
\subsection{Alignment}
Step-to-token: Rules/edit-distance; multi-aligner consensus; spot-checks.

\subsection{Causal Attribution}
Baselines (recon, nulls), linearity, hook sweeps, calibration. Stats sig. required~\citep{ScalingMonosemanticityBlog}.

\subsection{Validation}
Immediate: Purity (silhouette), clusters. Post: Ablations, replication, holdouts, audits.

\section{Evaluation Metrics}
- Purity: Top-k coherence, sparsity (target $\approx 30\%$).
- Selectivity: Top-5 activation variance (monosemanticity proxy).
- Causal: Task/probe shifts; feature ablation correlation ($r > 0.7$ target).
- Probe accuracy: Step prediction on CoT sequences.
- Cross-layer: Multi-model consistency.
- Transfer: Feature utility across models.

\section{Risks and Mitigations}
- Coverage: Phased scaling~\citep{SAE-Empirical}.
- Brittleness: Invariance tests~\citep{Li2025InterpretabilityIllusions}.
- Leakage: Ablations~\citep{SAE-Identifiability}.

\section{Resources (Optimized)}
50-100 RTX-equivalent hours, $<$100 GB; 1-2 weeks single GPU.

\section{Key Findings and Insights}

\textbf{Note (February 2026):} Phases 4, 5.1, 5.2, and 5.3 were executed with critical SAE training bugs and must be re-run before their results can be cited. Specifically: (1) \texttt{SAEConfig} defaulted to \texttt{use\_relu=False}, disabling L1 sparsity; (2) \texttt{normalize\_decoder()} was not called after optimizer steps, causing decoder collapse; (3) the decorrelation loss formula was incorrect (could produce negative loss). All sparsity percentages, transfer ratios, and causal correlations from those phases are unreliable. Phase 5.4 multilayer SAEs were retrained with the bugs fixed (\texttt{multilayer\_transfer\_v2/}) and are valid for use.

\subsection{Phase 4: Arithmetic Feature Probing — Mixed Results (Completed Feb 2026)}
SAEs (12$\times$ expansion, \texttt{l1\_penalty\_coeff=1e-4}) trained on GPT-2 medium residual streams, probed on 674 GSM8K arithmetic annotations (\texttt{<<expr=result>>} format):
\begin{itemize}
  \item \textbf{Experiment A (Ridge probe):} R$^2$=0.985 at layer 8 predicting $\log(|C|+1)$ from SAE features at result tokens. Features at operand tokens also strong (R$^2$=0.856). Layer 8 is the arithmetic peak across all positions.
  \item \textbf{Experiment B (Co-activation):} $\sim$23\,500 features ($\sim$1.9\% of 12\,288) show selectivity $>$0.1 for at least one arithmetic role. Computation-bridge features (active at both \texttt{=} and result) peak at layer 11.
  \item \textbf{Experiment C (Causal patching):} \textit{All} layers show negative $\Delta\log$-prob when top-128 probe features are swapped between examples (best: $-0.34$ at L0, worst: $-7.97$ at L9). Features are highly predictive but patching them in isolation disrupts rather than transfers arithmetic computation.
\end{itemize}
\textbf{Diagnosis:} SAE sparsity is $\sim$51\% (target $\approx30\%$) because L1 coefficient (1e-4) is too weak. At 51\% activation the features are polysemantic and co-dependent, so individual feature patches disrupt the distributed encoding rather than transferring it. This is the scenario warned against by~\citet{SAE-Identifiability}.

\textbf{Phase 4r — TopK SAEs + Subspace Steering (Completed Feb 2026):}
Retrained 24 GPT-2 medium 12$\times$ SAEs with hard TopK ($K=3686$, exactly 30\% sparsity).  Re-ran all three experiments:
\begin{itemize}
  \item \textbf{Experiment A:} R$^2$=0.977 at layer 7 (result tokens); instability at layers 5/7 from v1 eliminated.
  \item \textbf{Experiment B:} Feature taxonomy robust — computation-bridge 3.2\%, result-encoder 2.3\%, operand-tracker 1.9\%.
  \item \textbf{Experiment C (subspace steering):} \textit{Sign flipped.} Mean $\Delta\log$-prob $= +0.037$ across all layers; peak $+0.107$ at layer 22; 17/24 layers positive.  Arithmetic computation is causally concentrated in layers 12--23, with layer 22 as the strongest bottleneck.
\end{itemize}
\textbf{Conclusion:} Subspace steering (projecting $\Delta h = h_T - h_S$ onto the arithmetic decoder subspace) combined with 30\% sparsity resolves the causal failure from v1.  SAE features at upper layers causally encode arithmetic results, satisfying the falsification criterion for Phase 4.

\subsection{Phase 5: Feature Interpretation and Causal Steering (In Progress Feb 2026)}
Two experiments building directly on Phase 4r's TopK SAEs and the causal bottleneck at layer 22:
\begin{enumerate}
  \item \textbf{Feature auto-interpretation:} For each of the top-50 arithmetic features at causal layers (14--23), find the 10 max-activating token contexts in the 674-record GSM8K dataset.  Each feature receives a context card showing surrounding tokens, activation magnitude, arithmetic role (computation-bridge / result-encoder / operand-tracker), and expression/result values.  This converts statistical feature indices into human-readable descriptions.
  \item \textbf{Arithmetic steering:} Compute mean-difference vectors $\Delta f[L] = \overline{f_{\text{high-}C}}[L] - \overline{f_{\text{low-}C}}[L]$ in SAE feature space for each layer.  At test time, encode the residual, add $\alpha \Delta f[L]$, decode and inject.  Sweep $\alpha \in \{0.5, 1, 2, 4\}$ across all layers to find the (layer, $\alpha$) pair that most reliably shifts arithmetic predictions.  Success criterion: positive mean $\Delta\log$-prob toward high-magnitude results at the predicted best layer (22).
\end{enumerate}
Both experiments target GPT-2 medium first; extension to Phi-2, Gemma-2B, and Pythia-1.4B follows in Phase 6.

\subsection{Phase 3: Controlled CoT — Mixed Results (Completed Feb 2026)}
Linear probes trained on SAE latents from GPT-2 small (layer 6) on GSM8K reasoning steps:
\begin{itemize}
  \item \textbf{Equation token detection:} 83--97\% per-class accuracy across expansion factors 4x--20x. SAE features at layer 6 encode arithmetic computation tokens.
  \item \textbf{Reasoning/comparison connectives:} 0\% per-class accuracy across all expansions. Words such as ``therefore,'' ``so,'' ``more than'' are not linearly decodable from any single layer.
  \item \textbf{Overall:} No expansion factor beats the majority-class baseline (91.8\%). Single-layer linear probes are insufficient for multi-step reasoning structure.
\end{itemize}
\textbf{Implication:} Reasoning is not localized to a single layer. This motivates the multi-layer analysis approach of Phase 5.4, where SAE features are examined across all layers simultaneously.

\subsection{Phase 4B: Feature Monosemanticity (NEEDS REDO)}
Previous claim of consistent $31.7\%$ sparsity across all models is invalid (SAEs trained with \texttt{use\_relu=False}). Must be re-run with corrected training.

\subsection{Phase 5: Causal Analysis (NEEDS REDO)}
Previous selectivity--importance correlations ($r=0.754$ for Phi-2, etc.) and transfer ratios (0.995--1.000) are unreliable. Must be re-run after Phase 4 SAEs are retrained.

\subsection{Phase 5.2: Feature Naming (NEEDS REDO)}
Previous feature descriptions were generated from degenerate SAEs. Must be re-run with corrected SAE checkpoints.

\section{Extensions and Future Work}
\subsection{Phase 5.3: Feature Transfer Analysis (NEEDS REDO)}
Previous transfer ratios of $0.995$--$1.000$ are artifacts of degenerate SAEs (no ReLU gating, decoder collapse). Near-perfect reconstruction in a broken SAE does not indicate semantic universality. Must be re-run after Phase 4 SAEs are retrained with corrected training.

\subsection{Phase 5.4: Multi-Layer Feature Transfer Analysis (SAE CHECKPOINTS VALID; ANALYSIS PENDING)}
\textbf{Motivation:} Phase 5.3 examined only final-layer representations. Reasoning develops through network depths; different layers may show universal vs. specialized features. Phase 6 causal control requires layer-wise understanding.

\textbf{Approach:} Extend transfer analysis to mid-layer activations. Capture layer $\{4, 8, 12, 16, 20\}$ per model (4 models $\times$ 5 layers = 20 SAEs). Compute pairwise transfer for all combinations (400 transfer metrics). Generate layer universality heatmap.

\textbf{Research Questions:}
\begin{enumerate}
  \item Are early layers (4, 8, 12) universal across models?
  \item At what depth do models specialize?
  \item Bottleneck effect: Do models converge at certain layers?
  \item How do transfer metrics vary with network depth?
\end{enumerate}

\textbf{Expected Outcome:} Layer-wise universality map, identifying WHERE models diverge and optimal layers for Phase 6 feature steering. Integrated into Phase 6 design (layer-informed intervention strategies).

\subsection{Phase 6: Controllable Chain-of-Thought (Proposed)}
Extend to step-level causal testing: (1) Capture activations aligned with reasoning steps; (2) Test which features disrupt specific reasoning operations; (3) Measure step-wise importance (not global); (4) Steer reasoning via latent manipulation; (5) Identify universal reasoning primitives across architectures. 

\textbf{Phase 5.4 Integration:} Use multi-layer findings to target optimal layers for feature intervention, improving steering precision and interpretability.

Expected outcome: Deterministic control over multi-step reasoning without fine-tuning.

\subsection{Validation Criteria}

\textbf{Phase 5.4 Success Metrics:}
\begin{itemize}
  \item Multi-layer capture: All 20 layer activation files successfully generated
  \item SAE convergence: All 20 SAEs train to convergence (final loss $< 0.5$)
  \item Transfer matrix complete: 400 pairwise transfer metrics computed (no failures)
  \item Layer universality visible: Heatmap shows degradation pattern (validates architecture effects)
  \item No structural errors: All JSON outputs valid, no NaN/Inf values
\end{itemize}

\textbf{Phase 6 Validation Criteria:}
- Step-level feature importance: $r > 0.7$ between latent ablation and step quality
- Cross-model universality: Same features predict same steps (>70\% agreement)
- Steering efficacy: Controlled adjustment of reasoning length/quality via latent thresholds
- Generalization: Apply steering from one model, validate on another

\section{Conclusion}
This framework advances SAE interpretability for reasoning, with falsifiable phases and efficiencies. Phase 1 establishes that SAEs can faithfully recover ground-truth latent states from synthetic systems (R$^2$=0.967--0.999). Phase 3 reveals that single-layer probes detect arithmetic computation tokens (83--97\% per-class) but cannot decode reasoning connectives (0\% per-class) from any single layer --- motivating the multi-layer analysis of Phase 5.4. Phase 5.4 multilayer SAE checkpoints (98 SAEs across 4 models) have been retrained with corrected training and are valid for analysis. Phases 4, 5.1, 5.2, and 5.3 must be re-run with corrected SAE training before their results can be reported. Phase 6 toward controllable reasoning will be re-implemented once valid multi-layer SAE representations are established.

\bibliographystyle{plainnat}
\begin{thebibliography}{9}

\bibitem[Anthropic Mapping the Mind(2023)]{AnthropicMappingMind}
Anthropic.
\newblock Towards Monosemanticity: Decomposing Language Models With Dictionary Learning.
\newblock \url{https://transformer-circuits.pub/2023/monosemantic-features}, 2023.

\bibitem[Anon(2025a)]{SAE-Identifiability}
Anon.
\newblock Identifiable Sparse Autoencoders: Theory \& Methods.
\newblock arXiv:2506.15963, 2025.
\url{https://arxiv.org/abs/2506.15963}.

\bibitem[Anon(2025b)]{SAE-Random}
Anon.
\newblock Sparse Autoencoders Can Interpret Randomly Initialized Transformers.
\newblock arXiv:2501.17727, 2025.
\url{https://arxiv.org/abs/2501.17727}.

\bibitem[Anon(2025c)]{SAE-Domain}
Anon.
\newblock Resurrecting the Salmon: Rethinking Mechanistic Interpretability with Domain-Specific Sparse Autoencoders.
\newblock arXiv:2508.09363, 2025.

\bibitem[Anon(2024)]{SAE-Empirical}
Anon.
\newblock Evaluating Sparse Autoencoders for LLM Interpretability.
\newblock arXiv:2405.08366, 2024.
\url{https://arxiv.org/abs/2405.08366}.

\bibitem[Anon(2025d)]{SAE-Monosemantic}
Anon.
\newblock Sparse Autoencoders Learn Monosemantic Features in Vision-Language Models.
\newblock arXiv:2504.02821, 2025.
\url{https://arxiv.org/abs/2504.02821}.

\bibitem[Anthropic(2024)]{ScalingMonosemanticityBlog}
Anthropic.
\newblock Scaling Monosemanticity: Extracting Interpretable Features from Claude 3 Sonnet.
\newblock \url{https://transformer-circuits.pub/2024/scaling-monosemanticity/}, 2024.

\bibitem[Shu et~al.(2025)]{Shu2025SurveySAE}
Shu~et~al.
\newblock Sparse Autoencoders Survey in Mechanistic Interpretability.
\newblock 2025.

\bibitem[Li(2025)]{Li2025InterpretabilityIllusions}
Li.
\newblock Interpretability Illusions.
\newblock 2025.

\bibitem[Zhu(2025)]{Zhu2025AbsTopK}
Zhu~et~al.
\newblock AbsTopK and SAE Limitations.
\newblock 2025.

\end{thebibliography}

\end{document}